---
title: "The preprocessHighResMS User's Guide"
author: Silke Szymczak
package: preprocessHighResMS
abstract: >
  A comprehensive guide to using the preprocessHighresMS package for 
  preprocessing metabolomics data from direct infusion high resolution mass 
  spectrometry.
vignette: >
  %\VignetteIndexEntry{preprocessHighResMS User's Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document:
    toc_float: true
---

# Introduction

The preprocessHighResMS package provides methods to preprocess metabolomics 
data that are generated without a chromatography step (i.e. by direct infusion)
on high resolution mass spectrometers such as Orbitrap or Fourier-transform ion
cyclotron resonance mass spectrometry (FT-ICR-MS).

The package provides functions to define features based on chemical formulas 
and possible adducts, to extract the corresponding feature intensities from the
spectra, to filter features based on peak detection characteristics and to 
assign features to the most probable metabolite. Many of the functions 
use methods already implemented in different CRAN and Bioconductor packages.

The input are mass spectra in profile mode and stored in the mzML format. The 
tool msConvert from ProteoWizard 
(<http://proteowizard.sourceforge.net/tools.shtml>) can be used to convert 
the vendor specific files (e.g. .raw or .d) to the open, XML based data 
format mzML which is a community standard for mass spectrometry data. 

The package stores the feature and metabolite level information as a 
`r Biocpkg("SummarizedExperiment")` object which is a standard data structure 
for omics data sets provided by Bioconductor.

<!-- ## Citation -->

## Dependencies

This document has the following dependencies:

```{r dependencies, warning=FALSE, message=FALSE}
library(devtools)
library(preprocessHighResMS)
library(enviPat)
library(SummarizedExperiment)
library(ggpubr)
```

For the following analyses a data directory needs to be defined.

```{r}
data.dir = file.path(tempdir(), "example")
dir.create(data.dir)

```

## Example data

This vignette demonstrates the use of the package using example data from the
study MTBLS162 available in the MetaboLights database 
(<https://www.ebi.ac.uk/metabolights/MTBLS162/descriptors>). In this lipidomics 
study plasma samples from patients with severe insulin resistance caused by 
either loss-of-function mutations in the insulin receptor gene (INSR) or 
congenital generalised or partial lipodystrophy (LD) were compared with healthy 
controls. In addition, several blank and pooled quality control (QC) samples 
were included. 

Information about each sample is downloaded from the MetaboLights website and 
the relevant columns for the analysis are extracted.

```{r}
## define destination of phenotype file
pheno.file = file.path(data.dir, "info_pheno.txt")

## download file from MetaboLights
download.file(url = file.path("https://www.ebi.ac.uk/metabolights/ws/studies",
                              "MTBLS162/download",
                              "jFl7bH6WBo?file=s_Lypodystrophy.txt"),
              destfile = pheno.file)

## load phenotype information
pheno.all = read.table(file = pheno.file,
                       header = TRUE, as.is = TRUE, 
                       na.strings = c("N/A", "NCIT:Missing"))
pheno = pheno.all[, c("Sample.Name", 
                      "Source.Name", 
                      "Factor.Value.Disease.State.")]
colnames(pheno) = c("id", "source", "disease")
rownames(pheno) = pheno$id
head(pheno)

```

Lipids were measured with a nanoelectrospray method interfaced to 
the Thermo Exactive Orbitrap from Thermo Scientific.
The .raw files have been converted to mzML files using the msConvert tool with 
the following command:

```{bash, eval=FALSE}
msconvert.exe *.raw -o <output.dir> -e .mzML --mzML

```

The mzML files are available at BioStudies 
(https://www.ebi.ac.uk/biostudies/studies/S-BSST400)
and are downloaded into a subdirectory of `data.dir` called `spectrum`. Note 
that for one sample the mzML file could not be generated.

```{r}
spectrum.dir = file.path(data.dir, "spectrum")
dir.create(spectrum.dir)

## remove sample without mzML file
pheno = subset(pheno, id != "20121015b__002__12__ME")

## download
for (s in pheno$id) {
    download.file(url = file.path(
        "https://www.ebi.ac.uk/biostudies/files/S-BSST400",
        paste0(s, ".mzML")),
        destfile = file.path(spectrum.dir,
                             paste0(s, ".mzML")),
        quiet = TRUE)
}

```

# Defining features

Due to the high resolution of the used mass spectrometer a peak should be 
caused by a single feature, i.e. a specific isotope of an adduct of one of the 
metabolites that is contained in the sample. Thus, instead of starting with 
peak detection and assigning possible metabolites later in the annotation step,
this package starts with the metabolites that should theoretically be 
measurable in the samples. Here in this example several internal standards will
be used as well as some metabolites that show intensity differerences between 
controls and LD or INSR.

```{r}
## chemical formulas of standards
chem.forms.standards = c("C44H92NO6P",
                         "C45H94NO6P",
                         "C40H81N2O6P",
                         "C11H22O2",
                         "C39H74O6",
                         "C26H51NO3")

## chemical formulas of interesting metabolites
chem.forms.lipids = c("C41H81N2O6P",
                      "C41H83N2O6P",
                      "C44H80NO8P",
                      "C45H76O2",
                      "C55H100O6",
                      "C55H102O6")

```

The function `chem_formula_2_adducts()` can be used to calculate possible 
features. These features include different isotopes of specified adducts of the
chemical formulas. Here the adducts "M+H" and "M+NH4" are used and isotopes with
probabilities > 50% relative to the most intense isotope are included.
Before using this function it might be necessary to convert the chemical 
formulas which can be performed using the function `correct_chem_formula()`.

```{r}
## load isotope information
data("isotopes")

## correct chemical formulas
chem.forms.standards = correct_chem_formula(chem.forms = chem.forms.standards,
                                            isotopes = isotopes)

## slight modifications (e.g. P1 instead of P)
chem.forms.standards

chem.forms.lipids = correct_chem_formula(chem.forms = chem.forms.lipids,
                                            isotopes = isotopes)

chem.forms = c(chem.forms.standards,
               chem.forms.lipids)


## load adduct information
data("adducts")

## get possible features (M + adduct and different isotopes)
info.features = chem_formula_2_adducts(chem.forms = chem.forms,
                                       adduct.names = c("M+H", "M+NH4"),
                                       adducts = adducts,
                                       isotopes = isotopes,
                                       threshold = 50)
head(info.features)
nrow(info.features)
```

From the `r length(chem.forms)` chemical formulas `r nrow(info.features)` 
possible features have been calculated.

# Extract feature information

For each spectrum file intensities are extracted for each theoretical
feature using the function `extract_feature_intensity()`. The files with sample
specific feature information are stored in a subdirectory of `data.dir` called 
`features`.

```{r}
## create directory
feature.dir = file.path(data.dir, "features")
dir.create(feature.dir)

spectrum.files = dir(path = spectrum.dir,
                     full.names = TRUE)
for (f in spectrum.files) {
    extract_feature_intensity(spectrum.file = f,
                              scanrange = c(40, 50),
                              info.features = info.features,
                              ppm = 20,
                              res.dir = feature.dir,
                              verbose = FALSE)
}


```

A SummarizedExperiment object is created by combining the feature information
across samples. More information about this type of object is available at https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html.
Some of the features are removed if any of the following characteristics is 
fullfilled:

- missing mz values or intensities for all samples

- constant intensity across all samples 

```{r}
feature.files = dir(feature.dir,
                    full.names = TRUE)

se.all = combine_feature_intensities(feature.files = feature.files,
                                     pheno = pheno,
                                     verbose = TRUE)
```

In this example, none of the features is removed.

The SummarizedExperiment object contains information about the samples in the
`colData` slot. It would be possible to store information about the features in 
the `rowData` slot. The `assays` slot holds a list with several matrices with 
the features in the rows and the samples in the columns. In addition to the 
intensities (original and log-transformed), information about the detected peak
and the selected mz value are stored. 

```{r}
print(se.all)
```

# Filtering

While the previous filtering step is relevant for all studies, the following
criteria and thresholds need to be defined specifically for each study.

In the first step the relative frequency of detected peaks per sample is 
plotted and one sample is removed since it has an extremely low frequency of 
peaks.

```{r}
## frequency of peaks per sample
freq.peaks.sample = apply(assays(se.all)$peak.detection, 2, sum, na.rm = TRUE) /
    nrow(se.all)

dat.plot =  data.frame(freq.peaks.sample, 
                       source = se.all$source)
ggboxplot(dat.plot,
          x = "source",
          y = "freq.peaks.sample",
          add = "dotplot")

## remove sample with extremely low number of peaks
ind.rm = which(freq.peaks.sample < 0.1)
se.all = se.all[, -ind.rm]

```

Using the additional samples available in the study, features are kept if they
are detected in all of the QC samples but only in some of the blank 
samples.

After this QC step, the QC and blank samples are removed.

Furthermore, features with at least one missing value are removed. In contrast 
to the standard preprocessing approach, here missing values for a specific 
combination of feature and sample occur only if no or less than five intensities
were measured around the theoretical mz value of the feature in that sample.

Finally, features are removed if they have identical mz values for more than 
25% of the samples (same peak).

```{r}
## features detected in at least 50% of the quality control samples
ind.qc = which(se.all$source == "pooled quality control")
se.qc = se.all[, ind.qc]

peaks.feat.qc = apply(assays(se.qc)$peak.detection, 1, sum, na.rm = TRUE)
feat.detected.qc = names(peaks.feat.qc)[which(peaks.feat.qc == ncol(se.qc))]

## features detected in all blank samples
ind.blank = which(se.all$source == "blank sample")
se.blank = se.all[, ind.blank]

peaks.feat.blank = apply(assays(se.blank)$peak.detection, 1, sum, na.rm = TRUE)
feat.detected.blank = names(peaks.feat.blank)[which(peaks.feat.blank == 
                                                        ncol(se.blank))]

feat.use = setdiff(feat.detected.qc, feat.detected.blank)

## final data set
se = se.all[feat.use, -c(ind.qc, ind.blank)]

## check missing values in int
int = assays(se)$intensity
sum.na = apply(int, 1, function(x) {sum(is.na(x))})

## restrict to features without missing values
se = se[which(sum.na == 0), ]

## remove features with highly correlated or even identical mz values
## (corresponding to the same peak)
se = remove_features(se = se,
                     assay = "mz",
                     method = "identical.peaks")

print(se)
```

After filtering, only `r nrow(se)` features are left.

# Annotation

The measured features are assigned to the theoretical features using a Bayesian
approach that takes into account the distance between measured and theoretical
mz values as well as known relationships between adducts and isotopes.

First, all possible theoretical features within a defined window around each
measured feature are identified using the function `find_hits()` and the prior 
probabilities are estimated using the function `compute_prior_prob()`.

```{r}
dat = prepare_data_for_annotation(se = se)

hits.m = find_hits(info.features = info.features,
                   dat = dat,
                   ppm = 20)

prior.prob = compute_prior_prob(hits.m = hits.m,
                                info.features = info.features,
                                dat = dat,
                                ppm = 20)
max.pr = apply(prior.prob, 1, max)
```

One orphan isotope (= isotope for which the corresponding monoisotopic form was
not measured) is removed.

For illustration only a small number of features is considered in this example.
Thus, for each of the measured features the prior probability is 1 for the 
corresponding theoretical feature.
In a realistic analysis usually hundreds to thousands metabolites and the 
corresponding features are analysed so that prior probabilities can be much 
lower.

In order to incorporate additional information, connectivity matrices can be
calculated based on the function `generate_connectivity_matrix()` that contain
information about relationships between features e.g. based on different 
adducts of the same chemical component or isotopes of the same adduct.

Estimation of the posterior probabilities with the function 
`compute_posterior_prob()` usually takes some time and is skipped here.

```{r, eval = FALSE}
## not run
add.m = generate_connectivity_matrix(info.features = info.features,
                                     ids.use = colnames(hits.m),
                                     type = "adducts")
iso.m = generate_connectivity_matrix(info.features = info.features,
                                     ids.use = colnames(hits.m),
                                     type = "isotopes",
                                     ratios = FALSE)

set.seed(20200402)
post.prob.add.iso = compute_posterior_prob(prior.prob = prior.prob,
                                           dat = dat,
                                           add.m = add.m,
                                           iso.m = iso.m,
                                           delta.add = 0.1,
                                           delta.iso = 0.1)
max.po.a.i = apply(post.prob.add.iso, 1, max)
```

Based on either prior (here) or usually posterior probabilities, each measured
feature is assigned to the most probable theoretical feature or removed if the 
maximum probability is below a specified cutoff using the function 
`assign_features()`.

```{r}
info.assigned.use = assign_features(post.prob = prior.prob,
                                    dat = dat,
                                    cutoff.prob = 0.8)


```

# Metabolite level data

So far, all steps were based on the feature level. However, often several 
features belong to the same metabolite. For downstream statistical and 
bioinformatical analyses, the feature based metabolomics data is combined to 
metabolite data by summing up the intensitites of all features assigned to a 
particular metabolite in the function `summarize_metabolites()`.

```{r}
se.met = summarize_metabolites(info.assigned = info.assigned.use,
                               info.features = info.features,
                               se = se)

all(rownames(se.met) %in% chem.forms.lipids)
```

The resulting SummarizedExperiment object contains information about 
`r nrow(se)` metabolites that belong to the six lipids of interest.

For each of the metabolites an anlysis of variance (ANOVA) is performed 
followed by pairwise t-tests. The corresponding P-values are given in the 
boxplots.

```{r}
## define pairwise comparisons
my.comparisons = list( c("LD", "CON"), c("INSR", "CON"))

## plot log-transformed intensities
int = assays(se.met)$intensity.log2
int = int[sort(rownames(int)), ]

## boxplots with P-values using ggpubr
for (i in 1:nrow(int)) {
    dat.for.plot = na.omit(data.frame(intensity = int[i, ],
                                      group = se.met$disease,
                                      stringsAsFactors = FALSE))
    print(ggboxplot(dat.for.plot,
                    x = "group",
                    y = "intensity",
                    title = rownames(int)[i]) +
              stat_compare_means(method = "anova") +
              stat_compare_means(method = "t.test", 
                                 comparisons = my.comparisons))
}

```

All but one metabolite (C44H80N1O8P1) show significant differences between
LD and controls.

# Session info

```{r sessionInfo, echo=FALSE}
session_info()
```
